### 2016-07-03 store-consistency.tex
Previously uncommitted parts of ANSI ISO figure.

First, we observe that the semantics of a weakly-consistent data store
can be captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we can split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, resp.  $\I$ is now a
conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
trace invariant ($\I_c$) remains an invariant regardless of the
specific consistency and visibility features exposed by the underlying
data store, the store specific invariant ($\I_s$) changes from store
to store depending on the consistency level.  For an eventually
consistent store, $\I_s$ is simply \emph{true}. Stronger stores, such
as those that support strong consistent, have non-trivial definitions
for $\I_s$.

A strongly-consistent {\sc sc} store guarantees a total order on all
operations w.r.t. $\visZ$ consistent with their chronological order. A
straightforward $\I_s$ for this store is the \C{SC} property
formalized below:
\begin{smathpar}
\begin{array}{l}
  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all
transactions execute under \iso{Read Committed} isolation. Imagine a
sample execution where $\eta_1$'s transaction is not yet committed
when $\eta_2$ is generated. Letting $\eta_1$ be visible to $\eta_2$
violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
The only way to satisfy both invariants is to rule out all 
executions that interleave the operations of one transaction with the
other, thereby enforcing serializability.  In general, when $\I_s$ conflicts (but is not
inconsistent) with $\I_c$, the only way to enforce both invariant sets
is to restrict concurrency. Clearly, this is unacceptable since it
defeats the very purpose of supporting weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of
the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
    commitment}~\cite{bailishat} is often used in practice
  to capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
  visibility principle} to determine an acceptable weakening of $\I_s$
in case of a conflict with $\I_c$.  The principle requires the
weakened consistency guarantee ($\I_s'$) of the store to enforce all
visibility relationships imposed by the actual consistency guarantee
($\I_s$), unless enforcing such a relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
    {\sc sc} except in cases where they are in conflict with $\I_c$.
    Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
    Fig.~\ref{fig:txnimp} results in an operational semantics that
    describes many kinds of relational database-like data stores,
    including causally-consistent data stores, among
    others.\footnote{Details of this instantiation are relegated to
      the supplementary in the interest of space.}

>>>>

First, we observe that the consistency level of a data store can be
captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we can split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, respectively.  $\I$ is now a
conjunction: $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often
simplified to $\I \,=\, \I_s\wedge\I_c$).  While the program-specific
trace invariant ($\I_c$) is independent of the underlying data store,
the store specific invariant ($\I_s$) changes from store to store
depending on the consistency level.  For an eventually consistent
store, $\I_s$ is simply \emph{true}. Stronger stores, such as those
that support strong consistency, have non-trivial definitions for
$\I_s$.

A strongly-consistent {\sc sc} store guarantees a total order on all
operations w.r.t. $\visZ$ consistent with their chronological order. A
straightforward $\I_s$ for this store is the \C{SC} property
formalized below:
\begin{smathpar}
\begin{array}{l}
  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all
transactions execute under \iso{Read Committed} isolation. Imagine a
sample execution where $\eta_1$'s transaction is not yet committed
when $\eta_2$ is generated. Letting $\eta_1$ be visible to $\eta_2$
violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
The only way to satisfy both invariants is to rule out all executions
that interleave the operations of one transaction with the other,
thereby enforcing serializability.  In general, when $\I_s$ conflicts
(but is not inconsistent) with $\I_c$, the only way to enforce both is
to restrict concurrency. Clearly, this is unacceptable since it
defeats the very purpose of supporting weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of the
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is redefined
to \emph{true}, then the store makes $\eta_1$ visible to $\eta_2$ to
honor its consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailishat} is often used in practice to capture the
best-effort nature of {\sc sc}.}. We generalize this approach to any
$\I_s$ and $\I_c$ by defining a \emph{maximum visibility principle} to
determine an acceptable weakening of $\I_s$ in case of a conflict with
$\I_c$.  The principle requires the weakened consistency guarantee
($\I_s'$) of the store to enforce all visibility relationships imposed
by the actual consistency guarantee ($\I_s$), unless enforcing such a
relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
% GK : Please talk to me before changing the substance of the following paragraph.
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes {\sc sc} stores with on-demand weak isolation, such as
off-the-shelf relational databases. Similar approach can be employed
to capture the semantics of other kinds of stores, such as a causally
consistent ({\sc cc}) data store, as specific instantiations of our
operational semantics\footnote{Details of the {\sc cc} instantiation
are relegated to the supplementary in the interest of space.}.
